Dynamic Modeling, Parameter Estimation and Uncertainty Analysis in \proglang{R}

### 2016-10-03 article.tex
prediction profiles
Kleine updates

@@

To study bile clearance, experiments with hepatocytes or hepatocyte-like cell lines are performed \textit{in-vitro}. Cells in a Petri dish stick together to a monolayer of cells, forming bile canaliculi in between cells. They adhere to the plate and are surrounded by a buffer providing the cells with nutrients. To measure bile acids, here \textit{taurocholic acid} (TCA), radioactively labeled TCA is used.

@@

Each transportation process is modeled by mass-action kinetics. The implementation in \pkg{dMod} would be as follows:
\begin{CodeChunk}
	\begin{CodeInput}
# Add reactions
reactions <- NULL
reactions <- addReaction(reactions, "TCA_buffer", "TCA_cell",
			 rate = "import*TCA_buffer",
			 description = "Uptake")
reactions <- addReaction(reactions, "TCA_cell", "TCA_buffer",
			 rate = "export_sinus*TCA_cell",
			 description = "Sinusoidal export")
reactions <- addReaction(reactions, "TCA_cell", "TCA_cana",
			 rate = "export_cana*TCA_cell",
			 description = "Canalicular export")
reactions <- addReaction(reactions, "TCA_cana", "TCA_buffer",
			 rate = "reflux*TCA_cana",
			 description = "Reflux into the buffer")

@@

# Generate prediction function from ODE model
x <- Xs(mymodel, condition = NULL)
	\end{CodeInput}
\end{CodeChunk}
The reactions are collected in an \code{eqnlist} object. The \code{odemodel()} command composes the single reactions to an ODE system and auto-generates the \proglang{C} code which is used by the \pkg{deSolve} package to evaluate the ODE. Prediction functions generated by the \code{Xs()} command are wrapper functions for the call of \code{ode()}. The "s" in \code{Xs()} refers to "sensitivities", i.e the prediction function not only computes and returns the model prediction but sensitivities, too. The usage of the prediction function is illustrated by the following code chunk. Time points are defined between 0 and 50, numeric values are assigned to all model parameters. Finally the prediction function is called and the result, a prediction list, is plotted, shown in Figure~\ref{fig:prediction}A.

@@

# Get sensitivities for the dynamic parameters
out <- getDerivs(x(times, pars[4:7], fixed = pars[-(4:7)]))
plot(out)
\end{CodeInput}
\end{CodeChunk}

@@

g <- Y(observables, reactions, condition = NULL,
       compile = TRUE, modelname = "obsfn")

\end{CodeInput}
\end{CodeChunk}
The data we are going to simulate is for an efflux experiment, i.e. time point $t = 0$ reflects the distribution of TCA after uptake. Hence, we reset some of the parameter values. Initial values are set to steady state values. The incubation buffer is replaced by fresh buffer, meaning that \code{TCA_buffer} is set to 0. Finally, we assume that \code{s = 1000}. The prediction of the observables is made by concatenation of \code{g} and \code{x} via the concatenation operator \code{"*"} being defined in \pkg{dMod} for prediction, observation and parameter functions.
\begin{CodeChunk}
\begin{CodeInput}
# Reset parameter values
pars["TCA_cell"] <- 0.3846154
pars["TCA_cana"] <- 0.1538462
pars["TCA_buffer"] <- 0
pars["s"] <- 1e3

out <- (g*x)(times, pars, conditions = "standard")
\end{CodeInput}
\end{CodeChunk}
Finally, we obtain the simulated data by taking a subset of the prediction. At this point we make use of the "conditions" argument which allows to request a prediction for a vector of conditions, if provided by the prediction function\footnote{Both, \code{x} and \code{g} have been initialized with \code{condition = NULL} which means they are generic prediction functions and yield the same prediction for any of the requested conditions. The concept is further illustrated in Section~\ref{sec:conditions}.}. Data uncertainties $\sigma$ are derived by the Poisson nature of radioactive count experiments, i.e. $\sigma_x = \sqrt{x}$. To avoid division by 0, the minimal $\sigma$-value is set to 1. Random values are added to the predicted values to simulate observation noise. In the end, the \code{data.frame} is converted into a \code{datalist} object.
\begin{CodeChunk}
\begin{CodeInput}
# Simuate data
timesD <- c(0.1, 1, 3, 7, 11, 15, 20, 41)
datasheet <- subset(wide2long(out),
                    time %in% timesD & name %in% names(observables))
datasheet$sigma <- sqrt(datasheet$value + 1)
datasheet$value <- rnorm(nrow(datasheet), datasheet$value, datasheet$sigma)

@@

plot(out, data)
\end{CodeInput}
\end{CodeChunk}
Both, the simulated data and the model prediction from which the data is derived are shown in Figure~\ref{fig:observation}.
\begin{figure}[ht]
	\centering
	\includegraphics[width = \textwidth]{images/figure2}
	\caption{Model prediction of the internal states and observables. Simulated data is shown as dots with error bars.}
	\label{fig:observation}
\end{figure}

@@

\subsection{Objective function and model fitting}
For nomally distributed measurement noise, maximum-likelihood estimation is equivalent to least-squares estimation. A least-squares objective function can be generated by the \code{normL2()} command which requires a \code{datalist} object, in our case \code{data}, and a prediction function, in our case \code{g*x*p}.

@@

plotValues(myframe, value < 100)
plotPars(myframe, value < 100)
\end{CodeInput}
\end{CodeChunk}
Here, we have searched according to $\vec p_i = \vec p_0 + \Delta\vec p_i$ where $\vec p_0$ is the center, in our case \code{pouter}, $\Delta\vec p_i\sim N(0, \sigma^2)$ is a random parameter vector taken from a normal distribution, in our case $\sigma = 4$, and the index $i$ runs from 1 to \code{fits = 50}. The \code{mclapply()} command from the \pkg{parallel} package is used internally to run fits in parallel, here \code{cores = 4}. The result of \code{mstrust()} is a list of all returned values of \code{trust()}. To extract the final objective value, parameter values, convergence information and the number of interations, \code{as.parframe()} is used. The multi-start approach identifies four local optima, see Figure~\ref{fig:mstrust}, which yield almost the same objective value, Figure~\ref{fig:mstrust}A.
\begin{figure}[ht]
	\centering
	\includegraphics[width = \textwidth]{images/figure5}
	\caption{Result of multi-start fitting procedure. (A) Fits have been sorted by increasing least squares value. Four optima were found with almost identical objective value. (B) The parameter values for different optima are shown in different colors }
	\label{fig:mstrust}
\end{figure}
Despite the similar objective value, the optima are not close to each other in parameter space, as being illustrated by Figure~\ref{fig:mstrust}B. From the initial value parameters \code{TCA_cell} and \code{TCA_cana} it seems that they are exchangeable. Where \code{TCA_cell} is high, \code{TCA_cana} is low and vice versa. This is not surprising because the observed cellular TCA amount is the sum of both. A new experiment needs to be designed to distinguish one situation from another.

@@

In practice, the canaliculi only form a closed compartment if Ca$^{2+}$/Mg$^{2+}$ ions are present in the buffer. Therefore, if the experiment is repeated with Ca$^{2+}$/Mg$^{2+}$-free efflux buffer, the contents of the canaliculi escapes quickly into the buffer compartment and the observables \code{buffer} and \code{cellular} effectively reflect \code{TCA_buffer + TCA_cana} and \code{TCA_cell}. Mathematically, two experimental conditions which differ only by the \code{reflux} parameter need to be combined in one objective function.

@@

@@

data <- data + as.datalist(datasheet)
\end{CodeInput}
\end{CodeChunk}
Similar to objective functions and observation/prediction/parameter functions, as we will see, datalists can be combined by the \code{"+"} operator.

@@

Compared to the situation before we have gained the outer parameter \code{reflux_open}. Both transformations, "standard" and "open", evaluate the same vector of outer parameters meaning that they yield the same values for all but \code{reflux} and \code{reflux_open}. The prediction function \code{g*x} is generic in the sense that \code{condition = NULL} whereas the concatenation \code{g*x*p} has the conditions "standard" and "open", evaluating the identical function \code{g*x} on two parameter vectors.

@@

The parameters \code{export_sinus} and \code{reflux_open} exceed the 95\% confidence threshold only to one side. Given the data, the \code{export_sinus} parameter could equally be $-\infty$ (corresponding to an export rate of 0) without changing the likelihood significantly to the worse. A similar statement holds for the \code{reflux_open} parameter which could equally be $\infty$ meaning that we could assume instantaneous draining of the canaliculi for the \code{"open"} condition. The two parameters are practically non-identifiable and allow to reduce the model.

@@

The steady-state relation between \code{TCA_cana} and \code{TCA_cell} can be derived analytically from the ODE, Figure~\ref{fig:flowchart}A. It reads
\begin{equation}
	\verb!TCA_cana! = \verb!export_cana * TCA_cell / reflux!
	\label{}
\end{equation}
This relation can be explicitly used in a parameter transformation to express \code{TCA_cana} in terms of other parameters. The dimension of the parameter space is thereby reduced by one. The following implementation shows how we would use the existing transformation function \code{p} to generate an alternative transformation function \code{pSS} to replace \code{p} in the prediction function \code{g*x*p}.
\begin{CodeChunk}
\begin{CodeInput}
pSS <- NULL
trafos <- summary(p)
conditions <- names(trafos)
for (n in conditions) {

  equations <- trafos[[n]]$equations
  equations["TCA_cana"] <- "exp(export_cana)*exp(TCA_cell)/exp(reflux)"
  pSS <- pSS + P(equations, condition = n)

}
\end{CodeInput}
\end{CodeChunk}
We get all the information about the transformations from the \code{summary()} command. The equation for \code{TCA_cana} is substituted by our steady-state constraint. By the \code{"+"} operator, a new parameter transformation function \code{pSS} for our multiple conditions is constructed step-by-step.

Next, we want to implement the steady-state constraint by an implicit parameter transformation. Let $\dot x = f(x, p)$ be our dynamic system. Then, under certain conditions, we find a function $g(p)$ such that $f\big( g(p), p\big) = 0$ for all $p$ i.e.~$x_S = g(p)$ is a steady state of $f$. The parameter transformation we want to generate is the function $p \mapsto \big(p, g(p)\big)$. Here, the set of outer parameters is the set of the reaction rates  $p$ whereas the set of inner parameters contains these rates and the corresponding steady states as initial value parameters. The root of $f$ must be determined numerically which is implemented based on \code{multiroot()} from the \pkg{rootSolve} package.

@@

For the implicit parameter transformation we need the ODE which is obtained by the \code{as.eqnvec()} command from the reactions, an \code{eqnlist} object. The Jacobian of $f$ is rank-deficient because the system has a conserved quantity $c$, i.e.~\code{TCA_tot = TCA_buffer + TCA_cana + TCA_cell}. Replacing one element of $f$ by $c - \verb!TCA_tot!$, the rank of the Jacobian is completed, the condition for the local existence of the implicit function $g(p)$ is satisfied and the steady state is parameterized by $p$ and the additional parameter \code{TCA_tot}.

@@

for the dynamic state \code{TCA_cell} at time $t = 41$ under the "standard" condition. The uncertainty $\sigma = 0.002$ is set to a small value, i.e. below 1\% of the prediction value. The \code{datapointL2()} command returns an objective function which evaluates the model prediction\footnote{Several objective function combined by the \code{"+"} operator share the same environment. Thus, the prediction computed by the first objective function can be evaluated by all other functions to come.} and computes the least-squares function of the test datapoint, returning derivatives for the data-point parameter \code{d1}.

>>>>

The profile likelihood approach can be applied to compute confidence intervals for model predictions, too, see \citep{kreutz2012likelihood, hass2016fast}.

@@

To study bile clearance, experiments with hepatocytes or hepatocyte-like cell lines are performed \textit{in-vitro}. Cells in a Petri dish stick together to a monolayer of cells, forming bile canaliculi in between cells. They adhere to the plate and are surrounded by a buffer providing the cells with nutrients. To measure bile acid, here \textit{taurocholic acid} (TCA), it is labeled radioactively.

@@

Each transportation process is modeled by mass-action kinetics. A possible implementation in \pkg{dMod} is as follows:
\begin{CodeChunk}
	\begin{CodeInput}
# Add reactions
reactions <- NULL
reactions <- addReaction(reactions, "TCA_buffer", "TCA_cell",
			 rate = "import*TCA_buffer",
			 description = "Uptake")
reactions <- addReaction(reactions, "TCA_cell", "TCA_buffer",
			 rate = "export_sinus*TCA_cell",
			 description = "Sinusoidal export")
reactions <- addReaction(reactions, "TCA_cell", "TCA_cana",
			 rate = "export_cana*TCA_cell",
			 description = "Canalicular export")
reactions <- addReaction(reactions, "TCA_cana", "TCA_buffer",
			 rate = "reflux*TCA_cana",
			 description = "Reflux into the buffer")

@@

# Generate prediction function from ODE model
x <- Xs(mymodel, condition = NULL)
	\end{CodeInput}
\end{CodeChunk}
The reactions are collected in an \code{eqnlist} object. The \code{odemodel()} command composes the single reactions to an ODE system and auto-generates the \proglang{C} code which is used by the \pkg{deSolve} package to evaluate the ODE. Prediction functions generated by the \code{Xs()} command are wrapper functions for the call of \code{ode()}. The usage of the prediction function is illustrated by the following code chunk. Time points are defined between 0 and 50, numeric values are assigned to all model parameters. Finally the prediction function is called and both, the prediction and the sensitivities, are plotted, shown in Figure~\ref{fig:prediction}.

@@

# Get sensitivities for the rate parameters
out <- getDerivs(x(times, pars[4:7], fixed = pars[-(4:7)]))
plot(out)
\end{CodeInput}
\end{CodeChunk}

@@

g <- Y(observables, f = reactions, condition = NULL,
       compile = TRUE, modelname = "obsfn")

\end{CodeInput}
\end{CodeChunk}
Observation functions, linking internal states and observables states, can be used to simulate data. In the following, we will simulate the outcome of an \textit{efflux experiment}, i.e. time point $t = 0$ reflects the distribution of TCA after uptake, as shown in Figure~\ref{fig:prediction} after $t = 50$. Hence, we reset some of the parameter values. Initial values are set to steady state values. The incubation buffer is replaced by fresh buffer, meaning that \code{TCA_buffer} is set to 0. Finally, we assume that \code{s = 1000}. The prediction of the observables is made by concatenation of \code{g} and \code{x} via the concatenation operator \code{"*"} being defined in \pkg{dMod} for prediction, observation and parameter functions.
\begin{CodeChunk}
\begin{CodeInput}
# Reset parameter values
pars["TCA_cell"] <- 0.3846154
pars["TCA_cana"] <- 0.1538462
pars["TCA_buffer"] <- 0
pars["s"] <- 1e3

out <- (g*x)(times, pars, conditions = "standard")
\end{CodeInput}
\end{CodeChunk}
Since \code{g} and \code{x} have been generated as generic functions, i.e.~\code{condition = NULL}, we can assign the output to a condition of our choice, in thas case "standard".
Finally, we obtain the simulated data by taking a subset of the prediction. 
%At this point we make use of the "conditions" argument which allows to request a prediction for a vector of conditions, if provided by the prediction function\footnote{Both, \code{x} and \code{g} have been initialized with \code{condition = NULL} which means they are generic prediction functions and yield the same prediction for any of the requested conditions. The concept is further illustrated in Section~\ref{sec:conditions}.}.
Data uncertainties $\sigma$ are derived by the Poisson nature of radioactive count experiments, i.e. $\sigma_x = \sqrt{x}$. To avoid division by 0, the minimal $\sigma$-value is set to 1. Random values are added to the predicted values to simulate observation noise. In the end, the \code{data.frame} is converted into a \code{datalist} object.
\begin{CodeChunk}
\begin{CodeInput}
# Simuate data
timesD <- c(0.1, 1, 3, 7, 11, 15, 20, 41)
datasheet <- subset(wide2long(out),
                    time %in% timesD & name %in% names(observables))
datasheet$sigma <- sqrt(datasheet$value + 1)
datasheet$value <- rnorm(nrow(datasheet), datasheet$value, datasheet$sigma)

@@

plot(out, data)
\end{CodeInput}
\end{CodeChunk}
Both, the simulated data and the model prediction from which the data is derived are shown in Figure~\ref{fig:observation}. The data reflects a typical time course of an efflux experiment, showing decreasing cellular levels and increasing levels of TCA in the buffer.
\begin{figure}[ht]
	\centering
	\includegraphics[width = \textwidth]{images/figure2}
	\caption{Model prediction of the internal states and observables. Simulated data is shown as dots with error bars.}
	\label{fig:observation}
\end{figure}

@@

\subsection{Objective function and model fitting}
For normally distributed measurement noise, maximum-likelihood estimation is equivalent to least-squares estimation. A least-squares objective function can be generated by the \code{normL2()} command which requires a \code{datalist} object, in our case \code{data}, and a prediction function, in our case \code{g*x*p}.

@@

plotValues(myframe, value < 100)
plotPars(myframe, value < 100)
\end{CodeInput}
\end{CodeChunk}
Here, we have searched according to $\vec p_i = \vec p_0 + \Delta\vec p_i$ where $\vec p_0$ is the center, in our case \code{pouter}, $\Delta\vec p_i\sim N(0, \sigma^2)$ is a random parameter vector taken from a normal distribution, in our case $\sigma = 4$, and the index $i$ runs from 1 to \code{fits = 50}. The \code{mclapply()} command from the \pkg{parallel} package is used internally to run fits in parallel, here \code{cores = 4}. The result of \code{mstrust()} is a list of all returned values of \code{trust()}. To extract the final objective value, parameter values, convergence information and the number of interations, \code{as.parframe()} is used. The multi-start approach identifies four local optima, see Figure~\ref{fig:mstrust}, which yield almost the same objective value, Figure~\ref{fig:mstrust}A.
\begin{figure}[ht]
	\centering
	\includegraphics[width = \textwidth]{images/figure5}
	\caption{Result of multi-start fitting procedure. (A) Fits have been sorted by increasing least squares value. Four optima were found with almost identical objective value. (B) The parameter values for different optima are shown in different colors. (C) Each local optimum corresponds to a different model prediction, shown in different colors. The observed states are pracitically undistinguishable although the internal states show different behavior.}
	\label{fig:mstrust}
\end{figure}
Despite the similar objective value, the optima are not close to each other in parameter space, as being illustrated by Figure~\ref{fig:mstrust}B, and lead to different predictions, Figure~\ref{fig:mstrust}C. From the initial value parameters \code{TCA_cell} and \code{TCA_cana} it seems that they are exchangeable. Where \code{TCA_cell} is high, \code{TCA_cana} is low and vice versa. This is not surprising because the observed cellular TCA amount is the sum of both. A new experiment needs to be designed to distinguish one situation from another.

@@

In practice, the canaliculi only form a closed compartment if Ca$^{2+}$/Mg$^{2+}$ ions are present in the buffer. Therefore, if the experiment is repeated with Ca$^{2+}$/Mg$^{2+}$-free efflux buffer, the contents of the canaliculi escapes quickly into the buffer compartment. Under these conditions, the \code{buffer} measurement reflects what was formerly the total TCA content in buffer and canaliculi whereas the \code{cellular} measurement reflects what was fomerly the TCA content of the cells.
Mathematically, two experimental conditions which differ only by the \code{reflux} parameter need to be combined in one objective function.

@@

To simulate the new experimental condition, the \code{"reflux"} parameter is modified. The new data set is combined with the original data by the \code{"+"} operator.

@@

data <- data + as.datalist(datasheet)
\end{CodeInput}
\end{CodeChunk}

@@

Compared to the situation before we have gained the outer parameter \code{reflux_open}. Both transformations, "standard" and "open", evaluate the same vector of outer parameters meaning that they return the same values for all but the \code{reflux} parameter. The prediction function \code{g*x} is generic in the sense that \code{condition = NULL} whereas the concatenation \code{g*x*p} has the conditions "standard" and "open", evaluating the identical function \code{g*x} on two parameter vectors.

@@

The parameters \code{export_sinus} and \code{reflux_open} exceed the 95\% confidence threshold only to one side. Given the data, the \code{export_sinus} parameter could equally be $-\infty$ (corresponding to an export rate of 0) without changing the likelihood significantly to the worse. A similar statement holds for the \code{reflux_open} parameter which could equally be $\infty$ meaning that we could assume instantaneous draining of the canaliculi for the \code{"open"} condition. The two parameters are practically non-identifiable.

@@

The steady-state relation between \code{TCA_cana} and \code{TCA_cell} can be derived analytically from the ODE, Figure~\ref{fig:flowchart}A, setting the right-hand side of $\frac{\rm d}{ {\rm d}t} \verb!TCA_cana!$ to zero. It reads
\begin{equation}
	\verb!TCA_cana! = \verb!export_cana * TCA_cell / reflux!
	\label{}
\end{equation}
This relation can be explicitly used in a parameter transformation to express \code{TCA_cana} in terms of other parameters. The dimension of the parameter space is thereby reduced by one. The following implementation shows how we would use the existing transformation function \code{p} to generate an alternative transformation function \code{pSS} to replace \code{p} in the prediction function \code{g*x*p}.
\begin{CodeChunk}
\begin{CodeInput}
pSS <- NULL
trafos <- summary(p)
conditions <- names(trafos)

for (n in conditions) {
  equations <- trafos[[n]]$equations
  equations["TCA_cana"] <- "exp(export_cana)*exp(TCA_cell)/exp(reflux)"
  pSS <- pSS + P(equations, condition = n)
}
\end{CodeInput}
\end{CodeChunk}
We get all the information about the transformations from the \code{summary()} command. The equation for \code{TCA_cana} is substituted by our steady-state constraint. By the \code{"+"} operator, a new parameter transformation function \code{pSS} for our multiple conditions is constructed step-by-step.

Next, we want to implement the steady-state constraint by an \textit{implicit} parameter transformation, as opposed to the explicit transformation shown above. Let $\dot x = f(x, p)$ be our dynamic system. Then, under certain conditions, we find a function $g(p)$ such that $f\big( g(p), p\big) = 0$ for all $p$, i.e.~$x_S = g(p)$ is a steady state of $f$. The parameter transformation we want to generate is the function $p \mapsto \big(p, g(p)\big)$. Here, the set of outer parameters is the set of the reaction rates  $p$ whereas the set of inner parameters contains these rates and the corresponding steady states as initial value parameters. The root of $f$ must be determined numerically which is implemented based on \code{multiroot()} from the \pkg{rootSolve} package.

@@

For the implicit parameter transformation we need the ODE which is obtained by the command \code{as.eqnvec()} from the reactions, an \code{eqnlist} object. The Jacobian of $f$ is rank-deficient because the system has a conserved quantity $c = \verb!TCA_buffer + TCA_cana + TCA_cell!$ which is the total TCA amount. Replacing one element of $f$ by $c - \verb!TCA_tot!$, the rank of the Jacobian is completed, the condition for the local existence of the implicit function $g(p)$ is satisfied and the steady state is parameterized by $p$ and the additional parameter \code{TCA_tot}.

@@

for the dynamic state \code{TCA_cell} at time $t = 41$ under the "standard" condition. The uncertainty $\sigma = 0.002$ is set to a small value, i.e.~below 1\% of the prediction value. The \code{datapointL2()} command returns an objective function which evaluates the model prediction\footnote{Several objective function combined by the \code{"+"} operator share the same environment. Thus, the prediction computed by the first objective function can be evaluated by all other functions to come.} and computes the least-squares function of the test datapoint, returning derivatives for the data-point parameter \code{d1}.
